---
title: "STopover: Analysis of Visium dataset"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{STopover: Analysis of Visium dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  root.dir = "/home/nmadmin/DATA2/STopover"
)
```

```{r, eval = FALSE}
# Install packages
install.packages(c("Seurat","dplyr","tidyr","readr","pheatmap"))

# Install Bioconductor packages
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("clusterProfiler")
BiocManager::install("org.Mm.eg.db")
BiocManager::install("org.Hs.eg.db")

# Install CellDART
devtools::install_github("mexchy1000/CellDART", build_vignettes = T, force = T)

# Install STopover
devtools::install_github("bsungwoo/STopover", build_vignettes = T, force = T)
```

```{r install}
library(STopover)
library(CellDART)
library(Seurat)
library(dplyr)
library(tidyr)
library(readr)
library(pheatmap)
```

```{r}
# Make output folder
output_folder_name <- 'Results'
if (!file.exists(output_folder_name)){
  dir.create(output_folder_name)
}

fig_dir <- file.path(output_folder_name, 'figures')
if (!file.exists(fig_dir)){
  dir.create(fig_dir)
}
```

### 1. Load and preprocess single-cell datasets

##### Reference: Wu, S.Z. et al. A single-cell and spatially resolved atlas of human breast cancers. Nat Genet. 2021 Sep;53(9):1334-1347.

##### <https://singlecell.broadinstitute.org/single_cell/study/SCP1039>

```{r, eval = FALSE}
# Patient name for single-cell data
patient_list_sc <- c('CID4290A','CID4535','CID4465','CID44971')
tissue_type <- c("ER_1","ER_2","TNBC_1","TNBC_2")

# Load single-cell count matrix and metadata
sc_metadata <- read.csv('./Whole_miniatlas_meta.csv')
sc_metadata <- sc_metadata[-c(1),]
rownames(sc_metadata) <- sc_metadata$NAME
sc_metadata <- sc_metadata[-c(1,3)]
sc_metadata$nCount_RNA <- as.numeric(sc_metadata$nCount_RNA)
sc_metadata$nFeature_RNA <- as.numeric(sc_metadata$nFeature_RNA)

# Reverse log transform count matrix to get total normalized count
sc_data <- CreateSeuratObject(counts = expm1(expression_matrix))
# Add metadata and log-normalize
sc_data <- AddMetaData(sc_data, sc_metadata)
sc_data <- NormalizeData(sc_data, normalization.method = "LogNormalize", scale.factor = 1e4)
# Save single-cell Seurat object
readr::write_rds(sc_data, 'sc_data.rds')
```

### 2. Load and preprocess spatial datasets

##### <https://singlecell.broadinstitute.org/single_cell/study/SCP1039>

```{r, eval = FALSE}
# Patient name for spatial data
patient_list_sp <- c('CID4290','CID4535','CID4465','CID44971')
tissue_type <- c("ER_1","ER_2","TNBC_1","TNBC_2")

for (i in 1:length(patient_list_sp)){
  # Read count matrix
  breast_data <- Seurat::ReadMtx(paste0('./filtered_count_matrices/',patient_list_sp[i],
                                       '_filtered_count_matrix/matrix.mtx.gz'),
                                paste0('./filtered_count_matrices/',patient_list_sp[i],
                                       '_filtered_count_matrix/barcodes.tsv.gz'),
                                paste0('./filtered_count_matrices/',patient_list_sp[i],
                                       '_filtered_count_matrix/features.tsv.gz'),
                                feature.column = 1)
  breast_data <- Seurat::CreateSeuratObject(counts = breast_data, assay = "Spatial")
  # Read Visium images and coordinates
  breast_image <- Seurat::Read10X_Image(paste0('./spatial/',
                                               patient_list_sp[i],'_spatial/'))
  breast_image <- breast_image[Seurat::Cells(x = breast_data)]
  DefaultAssay(object = breast_image) <- 'Spatial'
  breast_data[[patient_list_sp[i]]] <- breast_image
  breast_data$orig.ident <- patient_list_sp[i]
  
  # Add metadata to the Seurat spatial data
  metadata <- read.csv(paste0('./metadata/',patient_list_sp[i],
                              '_metadata.csv'), header = T)
  rownames(metadata) <- metadata$X
  metadata <- metadata[,-(1)]
  breast_data <- AddMetaData(breast_data, metadata)
  breast_data$orig.ident <- factor(breast_data$orig.ident, levels=patient_list_sp[i])
  
  # Normalize the data and save
  breast_data <- NormalizeData(breast_data,
                               normalization.method = "LogNormalize",
                               scale.factor = 1e4)
  readr::write_rds(breast_data, file.path(output_folder_name, paste0('sp_data_',patient_list_sp[i],'.rds')))
}
```

### 3. Run CellDART analysis: predict spatial cell fraction

```{r, eval = FALSE}
# Load single-cell data (if needed)
sc_data <- readr::read_rds('sc_data.rds')

for (i in 1:length(patient_list_sp)){
  # Load spatial data (if needed)
  sp_data <- readr::read_rds(paste0('sp_data_',patient_list_sp[i],'.rds'))
  
  # Subset single-cell data to contain only the data obtained from the same patient
  sc_data_sel <- sc_data
  Idents(sc_data_sel) <- sc_data_sel$Patient
  sc_data_sel <- subset(sc_data_sel, idents = patient_list_sc[i])
  
  # Set the pseudospot number as the 5 times the number of spatial spots
  npseudo <- 5*dim(sp_data)[2]
  sp_data_cellf <- CellDART::pred_cellf_celldart(sp_data=sp_data,
                                                 sc_data=sc_data_sel,
                                                 outdir = output_folder_name,
                                                 env.select = "conda",
                                                 env.name = "CellDART",
                                                 gpu = TRUE,
                                                 metadata_celltype = "celltype_major",
                                                 num_markers = 20,
                                                 seed_num = 0,
                                                 nmix = 10,
                                                 npseudo = 20000,
                                                 alpha = 0.6,
                                                 alpha_lr = 5,
                                                 emb_dim = 64,
                                                 batch_size = 512,
                                                 n_iterations = 3000,
                                                 init_train_epoch = 10)
  
  ## Save Seurat object with cell fraction
  readr::write_rds(sp_data_cellf, file.path(output_folder_name,
                                            paste0('sp_data_cellf_',patient_list_sp[i],'.rds')))
}
```

### 4. Merge the Seurat objects with predicted cell fraction

```{r, eval = FALSE}
# Load the spatial data with predicted cell fraction and save as a list
sp_data_list <- lapply(patient_list_sp, function(x) {
  readr::read_rds(file.path(output_folder_name, 
                            paste0('sp_data_cellf_',x,'.rds')))})

# Include the spots that are located inside of the tissue
for (idx in 1:length(sp_data_list)){
  in_tissue_spots <- intersect(rownames(sp_data_list[[idx]]@images[[1]]@coordinates),
                               colnames(sp_data_list[[idx]]))
  sp_data_sub <- subset(sp_data_list[[idx]], cell=in_tissue_spots)
  sp_data_list[[idx]] <- sp_data_sub
}

# Merge the spatial Seurat objects
sp_data_merge <- merge(sp_data_list[[1]], sp_data_list[2:4])
sp_data_merge_mod <- sp_data_merge

# Remove the '_cellf' from the metadata columns (represents predicted cell fraction)
colnames(sp_data_merge_mod@meta.data) <- sapply(colnames(sp_data_merge_mod@meta.data),
                                                function(x) strsplit(x, split='_cellf')[[1]])
# Fill in NA values with 0
sp_data_merge_mod@meta.data[is.na(sp_data_merge_mod@meta.data)] <- 0
# save the Seurat object
readr::write_rds(sp_data_merge_mod, file.path(output_folder_name, 'sp_data_merge_mod.rds'))
```

### 5. STopover analysis: extract cell-cell colocalization patterns

```{r, eval=FALSE, echo=FALSE}
sp_data_merge_mod <- readr::read_rds(file.path('/home/nmadmin/DATA2/STopover_test/Results','sp_data_merge_mod.rds'))
```

```{r, eval=FALSE}
# Load cell type names saved in the medatata of single-cell data
celltype_names <- c('Normal.Epithelial','T.cells','Cancer.Epithelial','PVL',
                    'Endothelial','CAFs','B.cells','Myeloid','Plasmablasts')

# Define cell type pairs as calculated between single-cell and spatial data
feat_pairs <- data.frame(A="Cancer.Epithelial", B=celltype_names)

# Fun Stopover analysis between cancer epithelial cells and other cell types
sp_data_merge_celltype <- STopover::topological_similarity(sp_data_merge_mod, 
                                                           feat_pairs,
                                                           use_lr_db = F,
                                                           lr_db_species = "human",
                                                           min_size = 20,
                                                           fwhm = 2.5,
                                                           thres_per = 30,
                                                           assay = "Spatial",
                                                           slot = "data",
                                                           lognorm = F)

readr::write_rds(sp_data_merge_celltype, 'sp_data_merge_celltype.rds')
```
```{r, echo=FALSE}
sp_data_merge_celltype <- readr::read_rds('/home/nmadmin/DATA2/STopover_test/Results/visium/sp_data_merge_celltype.rds')
```

### 6. Draw cell-cell colocalization heatmap

```{r}
tissue_type_ <- c("ER","ER","TNBC","TNBC")

# Load global overlap score calculated between cancer epithelial and other cell types
df <- sp_data_merge_celltype@misc %>% 
  dplyr::select(batch, Feat_2, J_comp) %>% dplyr::filter(Feat_2!="Cancer.Epithelial") %>%
  tidyr::pivot_wider(names_from = batch, values_from = J_comp) %>%
  tibble::column_to_rownames(var = "Feat_2")
# Change column names (add tissue subtype information)
colnames(df) <- paste0(colnames(df), ": ", tissue_type_)
# Fill NA values with 0
df[is.na(df)] = 0

# Color choices
grDevices::hcl.pals()
```

```{r, fig.width=8, fig.height=8}
# Draw heatmap with pheatmap
out <- pheatmap::pheatmap(t(df), cluster_rows=T, show_rownames=T,
                          cluster_cols=T, angle_col=45, fontsize=15,
                          color = grDevices::hcl.colors(100, "Inferno"), 
                          na_col = "grey70",
                          breaks = seq(0, 0.3, by = 0.01*0.3),
                          clustering_method = 'average',
                          clustering_distance_rows = "correlation",
                          clustering_distance_cols = "correlation")
```

### 7. Visualize colocalized domain bewteen the cell type pairs

#### Visualize all interacting locations for the cell type pair

```{r, fig.width=6, fig.height=5}
tissue_type <- c("ER","ER","TNBC","TNBC")
patient_list_sp <- c('CID4290','CID4535','CID4465','CID44971')
slide_name <- paste0(patient_list_sp, ": ", tissue_type)

STopover::vis_all_connected(sp_data_merge_celltype, 
                            feat_name_x = "Cancer.Epithelial", 
                            feat_name_y = "Myeloid",
                            slide_titles = slide_name,
                            title_fontsize = 10,
                            legend_fontsize = 11,
                            save = T, slide_ncol = 2, dot_size = 1.6,
                            save_path = fig_dir,
                            save_name_add = paste0("_celltype"),
                            fig_width = 4, fig_height = 4, dpi = 150)
```

#### Visualize the top 2 interaction locations for the cell type pair

```{r, fig.width=8, fig.height=3.5}
STopover::vis_jaccard_top_n_pair(sp_data_merge_celltype, 
                                 feat_name_x = "Cancer.Epithelial", 
                                 feat_name_y = "Myeloid",
                                 top_n = 2,
                                 slide_name = "CID4465",
                                 title_fontsize = 10,
                                 legend_fontsize = 12,
                                 save = T, slide_ncol = 2, dot_size = 1.6,
                                 save_path = fig_dir,
                                 save_name_add = paste0("_celltype"),
                                 fig_width = 9, fig_height = 9, dpi = 150)

```

### 8. STopover analysis: extract dominant spatial LR interaction

```{r, eval=FALSE}
sp_data_merge_lr <- STopover::topological_similarity(sp_data_merge_mod,
                                                     use_lr_db = T,
                                                     lr_db_species = "human",
                                                     min_size = 20,
                                                     fwhm = 2.5,
                                                     thres_per = 30,
                                                     assay = "Spatial",
                                                     slot = "data",
                                                     lognorm = F)
readr::write_rds(sp_data_merge_lr, 'sp_data_merge_lr.rds')
```
```{r, echo=FALSE}
sp_data_merge_lr <- readr::read_rds('/home/nmadmin/DATA2/STopover_test/Results/visium/sp_data_merge_lr.rds')
```

#### Find and visualize differential upregulated LR interaction in TNBC \> ER types

##### Visualizing upregulated LR interactions and their corresponding GO terms using a heatmap

```{r}
tissue_type <- list("ER","ER","TNBC","TNBC")
patient_list_sp <- c('CID4290','CID4535','CID4465','CID44971')

result_list <- STopover::vis_diff_inc_lr_pairs(sp_data_merge_lr, 
                                               ref_group=c('CID4290','CID4535'), 
                                               comp_group=c('CID4465','CID44971'),
                                               logFC_cutoff=1, J_comp_cutoff=0.2,
                                               go_species="human",
                                               ontology_cat="BP",
                                               padjust_method = "BH", padjust_cutoff=0.05,
                                               top_n = 10, heatmap_max=5,
                                               title_fontsize=12,
                                               xaxis_title_fontsize=10,
                                               yaxis_title_fontsize=10,
                                               angle_col=45,
                                               colorbar_palette="RdPu",
                                               legend_loc='right',legend_fontsize=10,
                                               save_plot=F, save_path='.',
                                               save_name='heatmap_GO_LR_int',
                                               fig_width=21, fig_height=10, dpi=150,
                                               return_results=T)
```

```{r}
df_diff_TNBC <- result_list[[1]]
plot_diff_TNBC <- result_list[[2]]
```

```{r}
# Extract the to upregulated LR pairs in TNBC compared to ER+ tissues
df_diff_TNBC
```

```{r, fig.width=8, fig.height=6}
# Plot heatmap
plot_diff_TNBC
```

#### Visualize all interacting locations for the top LR pair

```{r, fig.width=6, fig.height=5}
tissue_type <- c("ER","ER","TNBC","TNBC")
patient_list_sp <- c('CID4290','CID4535','CID4465','CID44971')
slide_name <- paste0(patient_list_sp, ": ", tissue_type)

STopover::vis_all_connected(sp_data_merge_lr, 
                            feat_name_x = "IL18", 
                            feat_name_y = "IL1RL2",
                            slide_titles = slide_name,
                            title_fontsize = 10,
                            legend_fontsize = 11,
                            save = T, slide_ncol = 2, dot_size = 1.6,
                            save_path = fig_dir,
                            save_name_add = paste0("_celltype"),
                            fig_width = 4, fig_height = 4, dpi = 150)
```

#### Visualize the top 2 interaction locations for the top LR pair

```{r, fig.width=8, fig.height=3.5}
STopover::vis_jaccard_top_n_pair(sp_data_merge_lr, 
                                 feat_name_x = "IL18", 
                                 feat_name_y = "IL1RL2",
                                 top_n = 2,
                                 slide_name = "CID4465",
                                 title_fontsize = 10,
                                 legend_fontsize = 12,
                                 save = T, slide_ncol = 2, dot_size = 1.6,
                                 save_path = fig_dir,
                                 save_name_add = paste0("_celltype"),
                                 fig_width = 9, fig_height = 9, dpi = 150)
```
